A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-28 sections/Metatheory.tex
Appendix and all pointers
S4.6 Abella and Proof Statistics

We have proven soundness and completeness of the non-overlapping declarative
system with respect to the overlapping one to establish their equivalence.
Thus the restrictions do not change the expressive power of the system.
Modification (2) is relatively easy to justify, with the derivation given above:
the rule is redundant and can be replaced by a combination of three other rules.
Modifications (1) and (3) require inversion lemmas for the rules that overlap.
Firstly, Rule $\mathtt{\forall L}$ overlaps with Rule $\mathtt{\forall R}$ for the judgment
$\Psi \vdash \all A \le \all[b]B$.
The following inversion lemma for Rule $\mathtt{\forall R}$ resolves the overlap:
\begin{lemma}[Invertibility of $\mathtt{\forall R}$]
If $\Psi \vdash A \le \all B$ then $\Psi, a \vdash A \le B$.
\end{lemma}
The lemma implies that preferring Rule $\mathtt{\forall R}$ does not affect the derivability of the judgment.
Therefore the restriction $B \neq \all[b]B'$ in $\mathtt{\forall L'}$ is valid.

Secondly, Rule $\mathtt{DeclSub}$ overlaps with both $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$.
We have proven two inversion lemmas for these overlaps:
\begin{lemma}[Invertibility of $\mathtt{Decl\forall I}$]
If $\Psi \vdash e \Lto \all A$ then $\Psi, a \vdash e \Lto A$.
\end{lemma}
\begin{lemma}[Invertibility of $\mathtt{Decl{\to}I}$]
If $\Psi \vdash \lam e \Lto A \to B$ then $\Psi, x:A \vdash e \Lto B$.
\end{lemma}
These lemmas express that applying the more specific rules, rather than the more general rule $\mathtt{DeclSub}$, does not reduce the expressive power.

@@

\begin{lemma}[Subsumption]
Given $\Psi' \le \Psi$:
\begin{enumerate}
    \item If $\Psi \vdash e \Lto A$ and $\Psi \vdash A \le A'$ then $\Psi' \vdash e \Lto A$;
    \item If $\Psi \vdash e \To A$ then there exists
        $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash e \To A$;
    \item If $\Psi \vdash \appInf{C}{e}{A}$ and $\Psi \vdash C' \le C$,
        then there exists $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash \appInf{C'}{e}{A'}$.
\end{enumerate}
\end{lemma}
This lemma expresses that any derivation in a context $\Psi$ has a corresponding derivation in any context
$\Psi'$ that it subsumes.

We have tried to follow DK's manual proof of this lemma,
but we discovered several problems in their reasoning that we have been able to address.
Fortunately we have found a different way to prove the lemma.
The details of this issue can be found in Appendix~\ref{}.\bruno{reference}

@@

\begin{lemma}[Prune Transfer for Instantiation]
If $(\Gm \Vdash \al \le A \to B) \sto (\Om \Vdash C \le A_1 \to B_1)$ and
$\|\Om\| \vdash C \le A_1 \to B_1$, then $\al\notin FV(A) \cup FV(B)$.
\end{lemma}

@@

Combining all three main results (soundness, completeness and termination), we conclude that the declarative system is decidable
by means of our algorithm.
\begin{corollary}[Decidability of Declarative Typing]
Given \emph{wf }$\Om$, it is decidable whether $\Om\redto\nil$ or not.
\end{corollary}
\subsection{Abella}

>>>>

We have proven soundness and completeness of the non-overlapping declarative
system with respect to the overlapping one to establish their equivalence.
Thus the restrictions do not change the expressive power of the system.
Modification (2) is relatively easy to justify, with the derivation given above:
the rule is redundant and can be replaced by a combination of three other rules.
Modifications (1) and (3) require inversion lemmas for the rules that overlap.
Firstly, Rule $\mathtt{\forall L}$ overlaps with Rule $\mathtt{\forall R}$ for the judgment
$\Psi \vdash \all A \le \all[b]B$.
The following inversion lemma for Rule $\mathtt{\forall R}$ resolves the overlap:
\begin{lemma}[Invertibility of $\mathtt{\forall R}$]\label{lem:inv_allR}
If $\Psi \vdash A \le \all B$ then $\Psi, a \vdash A \le B$.
\end{lemma}
The lemma implies that preferring Rule $\mathtt{\forall R}$ does not affect the derivability of the judgment.
Therefore the restriction $B \neq \all[b]B'$ in $\mathtt{\forall L'}$ is valid.

Secondly, Rule $\mathtt{DeclSub}$ overlaps with both $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$.
We have proven two inversion lemmas for these overlaps:
\begin{lemma}[Invertibility of $\mathtt{Decl\forall I}$]\label{lem:inv_chkAll}
If $\Psi \vdash e \Lto \all A$ then $\Psi, a \vdash e \Lto A$.
\end{lemma}
\begin{lemma}[Invertibility of $\mathtt{Decl{\to}I}$]\label{lem:inv_chkLam}
If $\Psi \vdash \lam e \Lto A \to B$ then $\Psi, x:A \vdash e \Lto B$.
\end{lemma}
These lemmas express that applying the more specific rules, rather than the more general rule $\mathtt{DeclSub}$, does not reduce the expressive power.

@@

\begin{lemma}[Subsumption]\label{lem:subsumption}
Given $\Psi' \le \Psi$:
\begin{enumerate}
    \item If $\Psi \vdash e \Lto A$ and $\Psi \vdash A \le A'$ then $\Psi' \vdash e \Lto A$;
    \item If $\Psi \vdash e \To A$ then there exists
        $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash e \To A$;
    \item If $\Psi \vdash \appInf{C}{e}{A}$ and $\Psi \vdash C' \le C$,
        then there exists $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash \appInf{C'}{e}{A'}$.
\end{enumerate}
\end{lemma}
This lemma expresses that any derivation in a context $\Psi$ has a corresponding derivation in any context
$\Psi'$ that it subsumes.

We have tried to follow DK's manual proof of this lemma,
but we discovered several problems in their reasoning that we have been able to address.
Fortunately we have found a different way to prove the lemma.
The details of this issue can be found in Appendix~\ref{appendix:subsumption}.

@@

\begin{lemma}[Prune Transfer for Instantiation]\label{lem:prune_inst}
If $(\Gm \Vdash \al \le A \to B) \sto (\Om \Vdash C \le A_1 \to B_1)$ and
$\|\Om\| \vdash C \le A_1 \to B_1$, then $\al\notin FV(A) \cup FV(B)$.
\end{lemma}

@@

Combining all three main results (soundness, completeness and termination), we conclude that the declarative system is decidable
by means of our algorithm.
\begin{corollary}[Decidability of Declarative Typing]
Given \emph{wf }$\Om$, it is decidable whether $\Om\redto\nil$ or not.
\end{corollary}

% ---------------------------------------------
\subsection{Abella and Proof Statistics}
We have chosen the Abella (v2.0.7-dev) proof assistant~\cite{AbellaDesc} to develop our formalization.
Abella is designed to help such formalizations for programming languages,
due to its built-in support for variable bindings and
the $\lambda$-tree syntax~\cite{}, i.e. a form of HOAS.
Nominal variables, or $\nabla$-quantified variables, are used as an unlimited name supply,
which supports explicit freshness control and substitutions.

Although Abella lacks packages, tactics and support on modules,
we enjoy its higher-order system and the ease of formalizing substitution-intensive relations,
which are widely used in our algorithm.

\paragraph{Statistics of the Proof}
The proof script consists of 7683 lines of code with a total of
51 definitions and 567 theorems.
Figure~\ref{fig:proof_statistics} briefly summarizes the contents in each file.
The files are linealy dependent due to the limitation of Abella.

\begin{figure}
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  309 & 56  & Basic data structures\\\hline
    typing.thm          &  238 & 5   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strenthening\\\hline
    alg.thm             &  655 & 72  & Basic algorithmic properties\\\hline
    trans.thm           &  628 & 53  & \makecell[l]{Transfer and its basic properties \\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  944 & 72  & \makecell[l]{Additional properties for declarative typing \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll}, \ref{lem:inv_chkLam}}\\\hline
    soundness.thm       & 1069 & 70  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & \makecell[l]{The $|\cdot|_\to$ measure; \\
                            Lemma~\ref{lem:prune_inst}}\\\hline
    \makecell{dcl.thm, \\completeness.thm} & 1458 & 73  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  830 & 42  & Lemma~\ref{lemma:inst:decidable} Instantiation Decidability\\\hline
    decidability.thm    &  885 & 50  & Lemma~\ref{lemma:inst:decidable}\\\hline
    Total               & 7683 & 567 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}
